\documentclass{ximera}
\title{Compound Interest and the Number $e$}

\newcommand{\pskip}{\vskip 0.1 in}

\begin{document}
\begin{abstract}
Compound interest and the number $e$.
\end{abstract}
\maketitle


\section{Compound Interest}
 
\emph{Compound interest is the eighth wonder of the world. He who understands it, earns it; he who doesn't, pays it.} - Albert Einstein

\pskip \pskip
 
\begin{example} \label{Exdf831e3mzx}
An investment of $\$1,000$ earns interest at the \emph{nominal} rate of $12\%$/yr, compounded quarterly.

Find a function 
\[
   B = f(t) \, , \, t\geq 0,
\]
that expresses the balance (in dollars) in terms of the number of years since the investement was made.

\begin{explanation}
\emph{Nominal} means \emph{in name only}, and originates from the Latin \emph{nominalis} (relating to a name). Other words with the same root are 

\begin{enumerate}
\item \emph{nombre} (Spanish for \emph{name}).

\item \emph{nominate}

\item \emph{denomination}
\end{enumerate}

An investement that earns interest at the nominal rate of $12\%$/yr, compounded quarterly, does \emph{not} increase by $12\%$ each year. Rather, it increases at the \emph{effective} rate of 
\[
   \frac{12\%/\text{ yr}}{4 \text{ quarters/year}} = 3\%/\text{quarter} .
\]
This means that every quarter (3 months) the balance increases by $3\%$ and the one-quarter growth factor is
\[
   g = \answer{1.03} .
\]
So the one-year growth factor is
\[
   g^{\answer 4} = \answer{1.03^4} = 1.12550881.
\]
So every year the balance increases by 
\[
     \answer{12.550881}    \% 
\]
and the effective interest rate is
\[
               \answer{12.550881}    \% / \text{yr} . 
\]

So every year the balance gets multiplied by $g^4$ and with an initial investment of $\$1,000$, function that gives the balance (in dollars) after $t$ years is
\begin{align*}
   B &= f(t)  \\
      &= 1000\left( \answer{g^4} \right)^t  \\
      &= 1000\left( 1.03^4 \right)^{\answer{t}} \\ 
     &= 1000(1.03)^{\answer{4t}}
\end{align*}

\end{explanation}
\end{example}

\begin{example} \label{ExLdferr3r3}
An investment of $\$1,000$ earns interest at the \emph{nominal} rate of $12\%$/yr, compounded monthly.

\begin{enumerate}
\item Find a function 
\[
   B = f(t) \, , \, t\geq 0,
\]
that expresses the balance (in dollars) in terms of the number of years since the investement was made.


\item Find the balance after 10 years.

\item Find the balance after 40 years.

\item When is the balance $\$1,000,000$?

\item How much interest does the investment earn in the 10th year? The 40th year?

\item How much interest does the investement earn in the one-year period that starts when the balance is $\$100,000$?

\item Find a one-year period during which the investment earns $\$1,000$. Do not use a calculator. Then use a calculator to approximate the beginning of the one year period.

Start by defining an unknown, with units in a complete sentence.

\end{enumerate}
\end{example}


\section{The Number $e$}

\begin{example}  \label{E9re3rfdmer3110}
An investment of $\$1$ earns interest at the \emph{nominal} rate of $100\%$/yr.

Find the one-year growth factor if interest is compounded
\begin{enumerate}
\item annually

\item quarterly

\item monthly

\item daily

\item $1000$ times per year

\item $10000$ times per year

Use the worksheet below to help.

\begin{onlineOnly}
    \begin{center}
\desmos{1mbeajby8bb}{900}{600}
\end{center}
\end{onlineOnly}

\href{https://www.desmos.com/calculator/mbeajby8bb}{163: Hanging Weight and Tension 2}

\end{enumerate}
\end{example}

\emph{The Takeaway:} The number $e\sim 2.71828...$ is the one-year growth factor for an investment earning a nominal rate of $100\%$/yr, compounded continuously.

\begin{example} \label{ExKdf3rdf}
An investment of $\$1000$ earns interest at the nominal rate of $100\%$/yr.

\begin{enumerate}
\item Find functions 
\[
   B = f(t) \, , \, t\geq 0,
\]
and 
\[
  B = h(t) \, , \, t\geq 0,
\]
that express the balance (in dollars) in terms of the number of years since the investement was made if interest is compounded
\begin{enumerate}
\item $10$ times per year ($f$)

\item continuously ($g$).
\end{enumerate} 

\item Use your functions to answer each of the following questions. Compare your answers.

\begin{enumerate}
\item Find the balance after $10$ years.

\item When is the balance $\$1,000,000$?

\item Find a two-year period during which the investment earns $\$1,000,000$ in interest.
\end{enumerate}
\end{enumerate}

\end{example}


\begin{example}  \label{ExPfl3r8r3f31}
An investment of $\$1000$ earns interest at the nominal rate of $25\%$/yr, compounded continuously.

\begin{enumerate}
\item Find the four-year growth factor.

\item Find the one-year growth factor.

\item Find a function 
\[
   B = f(t) \, , \, t\geq 0,
\]
that expresses the balance (in dollars) in terms of the number of years since the investement was made.

\item Find the balance after $40$ years and compare with your answer to Question (b) (i) of Example 4.

\item When is the balance $\$1,000,000$? Compare with your answer to Question (b) (ii) of Example 4.

\item Find an $8$-year period during which the investement earns $\$1,000,000$ in interest. Compare with your answer to Question (b) (iii) of Example 4.
\end{enumerate}


\end{example}


\end{document}