\documentclass{ximera}
\title{Transformations, Part 1}

\newcommand{\pskip}{\vskip 0.1 in}

\begin{document}
\begin{abstract}
Transforming the graphs of functions and relations.
\end{abstract}
\maketitle

\section{Horizontal Translations}

\begin{question} \label{QP9erm3}
\begin{enumerate}
\item The graph of the function $y=f(x)$ is just the one point  with coordinates $(4,1)$. Find the coordinates of all points (if any) on the graph of the function
\[
  g(x) = f(x-5).
\]

\item  The point with coordinates $(a,b)$ lies on the graph of the function $y=f(x)$. What are the coordinates of the corresponding point on the graph of the function
\[
    g(x) = f(x-5)?
\]

\item The graph of the function $y=f(x)$ is shown below. 

\begin{onlineOnly}
    \begin{center}
\desmos{c5di6pnyda}{900}{600}
\end{center}
\end{onlineOnly}

\href{https://www.desmos.com/calculator/c5di6pnyda}{141: Trans Translation}

\begin{enumerate}
\item Write the domain and range of the function $y=f(x)$ in set-builder notation.

\item Find the domain and range of the function
\[
  g(x) = f(x-5) .
\]

\item Graph the function $y=g(x)$ \emph{by hand}. Label the coordinates of the three points corresponding to the three labeled points above.

\item Describe the physical transformation that takes the graph of $y=f(x)$ to the graph of $y=g(x)$.
\end{enumerate}
\end{enumerate}
\end{question}

\begin{question} \label{Q9d9r3ce4e}
Repeat all parts of Question 1 for each of the following functions using the graph of $y=f(x)$ in that question.

\begin{enumerate}
\item $g(x) = f\left(  \frac{1}{2} x\right)$

\item $g(x) = f\left(  \frac{1}{2}\left(x+5\right)\right)$

\item $g(x) = f\left(  \frac{1}{2}\left(x+5\right)\right)$

\item $g(x) = f(-x)$

\item $g(x) = f(|x|)$ for the function $f(x) = \log_2 x$. Start by graphing the function $y=\log_2 x$ by hand. 
\end{enumerate}
\end{question}


\begin{question} \label{QK3f3fsf}
The graphs of the function $y=f(x)$ and $y=g(x)$ are shown below.

\begin{onlineOnly}
    \begin{center}
\desmos{po1qtxqaxe}{900}{600}
\end{center}
\end{onlineOnly}

\href{https://www.desmos.com/calculator/po1qtxqaxe}{141: Trans Translation 2}

\begin{enumerate}
\item The point with coordinates $(a,b)$ lies on the graph of $y=f(x)$. What are the coordinates of the corresponding point on the graph of $y=g(x)$?

\item Express $g$ in terms of $f$. Check your answer with the three pairs of points labeled above. 
\end{enumerate}

\end{question}

\begin{question} \label{QK3f3fs33f}
The graphs of the function $y=f(x)$ and $y=g(x)$ are shown below.

\begin{onlineOnly}
    \begin{center}
\desmos{jetifi2e8e}{900}{600}
\end{center}
\end{onlineOnly}

\href{https://www.desmos.com/calculator/jetifi2e8e}{141: Trans Translation 3}

\begin{enumerate}
\item The point with coordinates $(a,b)$ lies on the graph of $y=f(x)$. What are the coordinates of the corresponding point on the graph of $y=g(x)$?

\item Express $g$ in terms of $f$. Check your answer with the three pairs of points labeled above. 

\item Express $f$ in terms of $g$. Check your answer with the three pairs of points labeled above.
\end{enumerate}

\end{question}

\begin{question} \label{QK3f3fefe33f}
The graphs of the function $y=f(x)$ and $y=g(x)$ are shown below.

\begin{onlineOnly}
    \begin{center}
\desmos{qhts1sps06}{900}{600}
\end{center}
\end{onlineOnly}

\href{https://www.desmos.com/calculator/qhts1sps06}{141: Trans Translation 4}

\begin{enumerate}
\item The point with coordinates $(a,b)$ lies on the graph of $y=f(x)$. What are the coordinates of the corresponding point on the graph of $y=g(x)$?

\item Express $g$ in terms of $f$. Check your answer with the three pairs of points labeled above. 

\item Express $f$ in terms of $g$. Check your answer with the three pairs of points labeled above.
\end{enumerate}

\end{question}





\begin{question} \label{QMfmdf33}
The graph of the function $y=f(x)$ is just the one point  with coordinates $(-7,3)$. Find the coordinates of all points (if any) on the graphs of each function below. Explain your reasoning.

\begin{enumerate}
\item $h_1(x) = f \left( \frac{x}{4} \right)$

\item $h_2(x) = f(4x)$.

\item $h_3(x) = f(x-12)$.

\item $h_4(x) = f(x^2 - 6x)$

\item $h_5(x) = f(2^x)$

\item $h_6(x) = f(\log_2 x)$

\item $h_7(x) = f(\sqrt{x})$

\item $h_8(x) = f \left( \frac{3-7x}{x-3} \right)$

\item $h_9(x) = f(b + mx) \, , \, m\neq 0$

\item $h_{10}(t) = f\left( \left| t  \right| \right)$

\item $h_{11}(t) = f\left( \left|7 t +3 \right| \right)$
\end{enumerate}

\end{question}


\begin{question} \label{QmQnwenSW}

The point $(a,b)$ lies on the graph of the function $y=f(x)$. Find the coordinates of points that lie on the graphs of each of the following functions. Explain your reasoning from first principles.

\begin{enumerate}
\item $g_1(x) = f\left( \frac{1}{2}x \right)$

\item $g_2(x) = f(2x)$

\item $g_3(t) = f\left( \frac{23}{13}t \right)$

\item $g_4(x) = f(-x)$

\item $g_5(t) = f\left( -\frac{7}{12}t \right)$

\item $g_6(t) = f(t-12)$

\item $g_7(x) = f(x+5)$

\item $g_8(x) = f(3(x+5)$

\item $g_9(x) = f(-3x + 12)$

\item $g_{10}(x) = f\left(\frac{13}{7} \left( x-5 \right) \right)$

\item $g_{11}(x) = f(mx+c) \, , \, m\neq 0 $
\end{enumerate}
\end{question}

\begin{question} \label{QMENFE3d0}
The point $(a,b)$ lies on the graph of the function $y=f(x)$. What point lies on the graph of the function
\[
       h(t) = f(g(t))?
\]
Assume $g$ is one-to-one and that $b$ is in range of $g$. 

What can you say if $b$ is not in the range of $g$? What if $g$ is not one-to-one?
\end{question}


\section{Constant Functions}
\begin{question} \label{Q9f8fmer}

The graph of the function $y=f(x)$ is shown below.

\begin{onlineOnly}
    \begin{center}
\desmos{401gwb7ubc}{900}{600}
\end{center}
\end{onlineOnly}

\href{https://www.desmos.com/calculator/401gwb7ubc}{141: Hor Trans 1}

\begin{enumerate}

\item Write the domain and range of $f$ in set-builder notation.

\item Use set-builder notation to write the domain and range of each of the following functions. Then graph each function.

\begin{enumerate}

\item $g_1(x) = f(x-12)$

\item $g_2(t) = f(9t + 18)$

\item $g_3(t) = f(t^2)$

\item $g_4(t) = f(2^t)$
\end{enumerate}

\end{enumerate}
\end{question}


\section{Applications}

\begin{question} \label{QKdfm3re32}
The function
\[
       G = f(m) \, , \, 0\leq m \leq 200 ,
\]
expresses the number of gallons of gas in your car as a function of the trip-odometer reading (in miles) during a 200-mile trip.

Let 
\[
    L = h(s)
\]
be the function that expresses the number of liters of gas in your car as a function of the trip-odometer reading (\emph{in km}) during the trip.

We'll assume there are exactly $1.6$km in one mile and exactly $4$ liters in one gallon. We'll also assume that $f(150) = 4$.

\begin{enumerate}
\item Write the domain of the function $h$ in set-builder notation.

\item Find the coordinates of a point on the graph of the function $G=h(s)$.

\item If the point with coordinates $(a,b)$ lies on the graph of $G=f(m)$, what are the coordinates of the corresponding point on the graph of $L=h(s)$?

\item Express the function $h$ in terms of $f$. Use the correct input and the correct output.

\item If the point with coordinates $(a,b)$ lies on the graph of $L=h(s)$, what are the coordinates of the corresponding point on the graph of $G=f(m)$?

\item Express $f$ in terms of $h$. Use the correct input and the correct output.

\item Use the graph of the function $G=f(m)$ below to sketch (by hand) a graph of the function $L=h(s)$. Label the axes with the appropriate variable names and units.

\begin{onlineOnly}
    \begin{center}
\desmos{nh0w4bcvby}{900}{600}
\end{center}
\end{onlineOnly}

\href{https://www.desmos.com/calculator/nh0w4bcvby}{141: Gas Consumption Trans}
\end{enumerate}
\end{question}

\begin{question} \label{Q9d9r3mdf32}
The function $h=f(\theta)$, $0\leq \theta \leq 1$, expresses your height (in feet) in terms of the rotation angle of a ferrris wheel (measured in fractions of a revolution from the downward vertical). Its graph is shown below.

\begin{onlineOnly}
    \begin{center}
\desmos{olrqe90xht}{900}{600}
\end{center}
\end{onlineOnly}

\href{https://www.desmos.com/calculator/olrqe90xht}{141: Ferris Wheel 1}

The function
\[
   h = g(t) \, , \, 0.5 \leq t \leq 3 ,
\]
expresses your height (in feet) in terms of the number of minutes past noon as you ride the ferris wheel. Its graph is shown below.

\begin{onlineOnly}
    \begin{center}
\desmos{szsqtovplg}{900}{600}
\end{center}
\end{onlineOnly}

\href{https://www.desmos.com/calculator/szsqtovplg}{141: Ferris Wheel 2}

\begin{enumerate}
\item What is the radius of the ferris wheel? Explain your reasoning.

\item Graph the function
\[
       \theta = a(t) \, , \, 0.5 \leq t \leq 3 ,
\]
that expresses the rotation angle of the wheel (in revolutions) in terms of the number of minutes past noon. Label the axes with the appropriate variable names and units.

\item Find an expression for the function $\theta=a(t)$.

\item Express the function $g$ in terms of the function $f$. Use the appropriate variable names for the input and output.

\item Express the function $f$ in terms of the function $g$. Use the appropriate variable names for the input and output.
\end{enumerate}

\end{question}


\end{document}
