\documentclass{ximera}
\title{Some Problems for Review}

\newcommand{\pskip}{\vskip 0.1 in}

\begin{document}
\begin{abstract}
Review problems.
\end{abstract}
\maketitle

Here are some sample finals. They are \emph{not} necessarily indicative of the exact types of problems you will see on the final, nor are they meant to be an exhaustive list of all the types of problems. But they should give you some general idea of what to expect.

Neither calculators or notes will be permitted. Do \emph{not} use the quadratic formula.

\section{Sample 1}

\begin{question}  \label{Q34gt4r5t4t44}
\begin{enumerate}
\item Find an equation of the circle centered at the point $A(3,-5)$ that passes through the point $Q(-4,-1)$.

\item Explain the \emph{logic} behind your equation.
\end{enumerate}

\end{question}


\begin{question}  \label{Qdftr4tytuiiii}
Use the methods of \emph{our class} to find an equation of the circle through the points $A(2,1)$, $B(-3,4)$, and $C(-1,-3)$. Explain your reasoning. 

\end{question}


\begin{question}  \label{Qhjk88654g6}
The function 
\[
         T =  f(h) , 0\leq h \leq 24,
\]
expresses the temperature (in Fahrenheit degrees) in terms of the number of hours past midnight on a summer day in Seattle.

Translate each of the following questions into mathematics. Each translation must begin with a verb.
\begin{enumerate}
\item Find the temperature at 2:00pm.

\item When was the temperature $73^\circ$F?

\item When was the temperature the same as it was at 5:00am.

\item When was the temperature $5$ Fahrenheit degress less than it was as 4:00pm?

\item When was the temperature at least $5$ Fahrenheit degress less than it was as 4:00pm?

\item When is the temperature the same as it was four hours ago?

\item When is the temperature $3$ Fahrenheit degress less than it was four hours ago?
\end{enumerate}

\end{question}


\begin{question}  \label{Qsdfsdftiii}
Let 
\[
        f(t) = t^2 - 10t + 1 .
\]
Do \emph{not} rely on the quadratic formula in solving these equations. Write a concluding sentence with a solution set where appropriate.
\begin{enumerate}
\item Solve the equation $f(x) = 10$.

\item Solve the equation $f(t) = f(9)$.

\item Find all inputs to the function $f$ that are equal to their outputs.
\end{enumerate}
\end{question}


\begin{question} \label{Q34gfgryt55yt5t}
The function
\[
    h = f(t) = 242 - 8 (t-5)^2 \, , ,\, 0\leq t \leq 10.5
\]
expresses the height (in feet) of a rock launched vertically upward on the planet Krypton in terms of the number of seconds since it was launched.

\begin{enumerate}
\item Find an expression for the function
\[
 t = d(h)
\]
that gives the time when the ball is $h$ feet high and on its way down.

\item Express the domain and range of the function $d$ in set notation.

\item Explain the meaning of the composition $d(f(3))$.

\item Evaluate the compostion $d(f(3))$.

\item Simplify the composition
\[
    g(t) = d(f(t)).
\]

\item Express the domain and range of the function $g$ in set notation.

\item Graph the function $T=g(t)$.


\end{enumerate}
\end{question}

\begin{question} \label{Q5gbrghbnnmvcvhn}
The graph of the function $y=f(x)$ is shown below.


\begin{onlineOnly}
    \begin{center}
\desmos{got9sylex8}{900}{600}
\end{center}
\end{onlineOnly}


\href{https://www.desmos.com/calculator/got9sylex8}{141: Graph Problem 12}


\begin{enumerate}
\item Approximate the solution(s) to the equation 
\[
 f(f(x)) = 4.
\]
Write your answer as a solution set. Explain your reasoning.

\item Find, if possible, a real number $c$ so that the equation
\[
   f(f(x))=c
\]
has exactly three solutions. Explain your reasoning. Then approximate the three solutions if possible.

\item Suppose
\[
  f(x) = -1 + \frac{1}{12}(12x-x^2) \, , \, 0\leq x \leq 12 ,
\]
to find the exact solutions to the equations in (a) and (b). Write your final answers as a solution set.
\end{enumerate}

\end{question}

\begin{question} \label{Qgcgrt45r5r5}
Let
\[
   f(x) = \sqrt{100-x^2}.
\]

\begin{enumerate}
\item Express the domain and range of $f$ in set notation.

\item Graph the function $y=f(x)$ by hand.

\item Simplify the composition $g(x) = f(f(x))$.

\item Express the domain and range of $g$ in set notation.

\item Graph the function $y=g(x)$.

\end{enumerate}
\end{question}




\end{document}