\documentclass{ximera}
\title{Functions, CW3}

\newcommand{\pskip}{\vskip 0.1 in}

\begin{document}
\begin{abstract}
Using function notation to translate between English and mathematics; reading information from the graphs of functions.
\end{abstract}
\maketitle

\section{What is a Function?}


\begin{question}  \label{Qdghf4rr3gg4}
(a) During a non-stop 280-mile car trip from Seattle to Spokane would you expect the speed of your car to be a function of your distance from Seattle? Explain.

(b) During the same trip would you expect the distance from Seattle to be a function of your car's speed? Explain.
\end{question}

\begin{question}  \label{Qer435r343554}
(a) During an 80-mile round trip journey from Seattle to North Bend and back, would you expect the speed of your car to be a function of your distance from Seattle? Explain.

(b) During the same trip would you expect the speed of your car to be a function the odometer reading? Explain.
\end{question}


\begin{question}  \label{Q:4sdfdslk8}

Play the slider $u$ in the activity below to see the motion of a balloon during a 40-second interval. (ignore the little jig to the right at the bottom of the path and assume the balloon moves straight up and down). Use the animation to answer the following questions. Explain your reasoning for each.

(a) Is the speed of the balloon a function of the balloon's altitude during the 40 second interval?

(b) Is the speed of the balloon a function of altitude when the balloon is on its way down?

(c) Is the altitude a function of speed on the way down?

\begin{onlineOnly}
    \begin{center}
\desmos{amv52b9ljt}{900}{600}
\end{center}
\end{onlineOnly}

Desmos activity available at \href{https://www.desmos.com/calculator/amv52b9ljt}{151: Balloon}

\end{question}


\section{Translating Between Math and English}

\begin{question}  \label{QFun25}

The function
\[
   g = f(v) , 20 \leq v \leq 75 ,
\]
expresses the gas mileage (in miles/gal) of a car in terms of its speed (in miles/hour). Translate each of the following statements into everyday Engish or into mathematics.

\begin{enumerate}

\item Solve the equation $f(v) = 40$.

\item Evaluate $f(40)$.

\item Determine the speed(s) at which the car gets $32$ miles/gal.

\item At what speeds does the car get at most $32$ miles/gal?

\item Solve the equation $f(v)=f(40)$.

%\item Solve the inequality $f(v)\leq 38$.

\item Solve the inequality $f(v) geq f(40)$.

\item Solve the equation $f(v)=f(v-8) + 3$.

\end{enumerate}
\end{question}


\begin{question}  \label{QFun354}
Use the graph of the function 
\[
    g = f(v) , 20\leq v \leq 75 ,
\]
shown below and the sliders to find approximate answers to each of the following questions. Answer each mathematical question with mathematics (write the solutions to an equation or inequality as a set). Answer each English question with a complete sentence.
\pdfOnly{
Access Desmos interactives through the online version of this text at
 
\href{https://www.desmos.com/calculator/pkn5m5wfky}.
}
 
\begin{onlineOnly}
    \begin{center}
\desmos{mly7wxfxtj}{900}{600}
\end{center}
\end{onlineOnly}

\href{https://www.desmos.com/calculator/mly7wxfxtj}{141: Gas Mileage 70}

\begin{enumerate}

\item Evaluate $f(40)$. 

\item  At what speeds does the car get 35 miles/gal?

\item  At what speeds does the car get at least 35 miles/gal?

\item  At what speeds does the car get at most 35 miles/gal?

\item Solve the inequality $f(v)> 50$.

\item Solve the inequality $f(v)<50$.

\item  At what speeds is the gas mileage $25\%$ less than it is at a speed of 50 miles/hour?

\end{enumerate}

\end{question}


\begin{question} \label{QMDM3er3r3}

The function 
\[
    g = f(v) = 40-\frac{1}{25}\left(v-50\right)^{2} \, , \, 35\leq v \leq 70 ,
\]
expresses the gas mileage (in miles/gal) of a car in terms of its speed.

\begin{enumerate}

\item Use the graph of the function $f$ below to find approximate answers to the following questions. Answer each question with a complete sentence.

\item Then use algebra to find the exact answers. Again, answer each question with a complete sentence.

\end{enumerate}

\begin{onlineOnly}
    \begin{center}
\desmos{rmpah0rcez}{900}{600}
\end{center}
\end{onlineOnly}

\href{https://www.desmos.com/calculator/rmpah0rcez}{142: Gas Mileage 45}

\begin{enumerate}
\item At what speed(s) does the car get $35$ miles/gal? Find the exact speed(s) without a calculator. Then use a calculator to approximate the speed(s) to the nearest tenth of a mile/hour.

\item At what speed(s) is the gas mileage the same as it is at a speed of $38$ miles/hour. Do \emph{not} use a calculator.

\item At what speed(s) does increasing the speed by $10$ miles/hour leave the gas mileage unchanged. 

\item  At what speed(s) does decreasing the speed by $10$ miles/hour increase the gas mileage by $8$ miles/gal?
\end{enumerate}
\end{question}











\end{document}
