\documentclass{ximera}
\title{Centers Set, Part 1}


\newcommand{\pskip}{\vskip 0.1 in}

\begin{document}
\begin{abstract}
More questions about circles.
\end{abstract}
\maketitle


\pskip

\section{Distance from a Point to a Line}

\begin{question} \label{Qpefrp3r}
\begin{enumerate}
\item Find the distance from the point $(3,5)$ to the $x$-axis.

\item Find the distance from the point $(2,-8)$ to the $x$-axis.

\item Find an expression for the distance from the point $(a,b)$ to the $x$-axis.

\item Find an expression for the distance from the point $(a,b)$ to the line $y=3$.

\item Find an expression for the distance from the point $(a,b)$ to the line $x=-2$.
\end{enumerate}
\end{question}


\begin{question} \label{Qdf93r1000}

You'll learn about tangent lines to a curve in calculus, but for now we'll say that a line is tangent to a circle if it intersects the circle in exactly one point.


\begin{onlineOnly}
    \begin{center}
\desmos{aosglrotza}{900}{600}
\end{center}
\end{onlineOnly}

\href{https://www.desmos.com/calculator/aosglrotza}{141: Circles tangent to lines}

\begin{enumerate}

\item The circle centered at $(3,4)$ in the worksheet above has radius $r$. Drag the slider $r$ in Line 1 to make the circle tangent to the $x$-axis. 

\item Find an equation of the circle centered at the point $(3,4)$ that is tangent to the $x$-axis.

\item Find an equation of the circle centered at the point $(3,4)$ that is tangent to the line $x=-2$.

\item Find an equation of the circle centered at the point $(3,4)$ that is tangent to the line $x=-2$.

\item Find an equation of the circle in the second quadrant with radius $5$ that is tangent to both the $x$ and $y$-axes. Enter your equation in Line 2 to see if you are correct.

\item How many circles of radius $3$ are tangent to the lines $y=-2$ and $x=-5$? Find an equation of one such circle. Enter your equation in Line 3 to see if you are correct.

\end{enumerate}

\end{question}


\section{A Center Set}

\begin{question}  \label{QKdf3Dfer33}
This question is about finding equations of all circles through the point $A(2,3)$ that are tangent to the $x$-axis.
\begin{enumerate}

\item Is the point $(-1,4)$ the center of some circle through the point $A(2,3)$ that is tangent to the $x$-axis? Explain your reasoning. Draw a picture to help.

\item Drag the centers of the circles below so that the circles are tangent to the $x$-axis and pass through the point $A(2,3)$.

\begin{onlineOnly}
    \begin{center}
\desmos{ghthxwn4uk}{900}{600}
\end{center}
\end{onlineOnly}

\href{https://www.desmos.com/calculator/ghthxwn4uk}{141: Circles tangent to a line}

\item What do you notice about the centers of these circles? Make a guess.

\item Find necessary and sufficient conditions for the point $P$ with coordinates $(x,y)$ to be the center of a circle through the point $A(2,3)$ and tangent to the $x$-axis.

Start by completing this sentence in everyday English. Then translate your condition into an equation.

\pskip

\emph{A point $P$ with coordinates $(x,y)$ is the center of a circle through the point $A(2,3)$ and tangent to the $x$-axis if and only if ...}

\pskip

\item Enter your equation from part (d) in Line 2 of the worksheet above.

\item Find equations of all circles with radius $7$ through the point $A(2,3)$ and tangent to the $x$-axis. Enter these equations in Lines 4 and 5 above to see if you are correct.

\item Find equations of all circles with radius $r$ through the point $A(2,3)$ and tangent to the $x$-axis. Enter these equations in Lines 6 and 7 above. Then drag the slider $r$ in Line 8 to see if you are correct.
 
\item For what value of $r$ are there exactly 

\begin{enumerate}
\item zero

\item one

\item two
\end{enumerate}
circles with with radius $r$ through the point $A(2,3)$ and tangent to the $x$-axis.

\item Find an equation of the smallest circle through the point $(2,3$ that is tangent to the $x$-axis.
\end{enumerate}
\end{question}
\end{document}