\documentclass{ximera}
\title{Exponential Growth}

\newcommand{\pskip}{\vskip 0.1 in}

\begin{document}
\begin{abstract}
Exponential functions and their inverses.
\end{abstract}
\maketitle

\begin{exploration}  \label{Edftrg455}

\begin{enumerate}
\item The animation below shows an expanding circle with a radius that grows exponentially. Desribe what you see. Be as precise as possible. Use the animation only. Do \emph{not} open any folders.
\begin{freeResponse}
\end{freeResponse}

\item A friend says that a population growing exponentially grows very rapidly. How would you respond?
\begin{freeResponse}
\end{freeResponse}

\end{enumerate}


\begin{onlineOnly}
    \begin{center}
\desmos{8k0cwumzso}{900}{600}
\end{center}
\end{onlineOnly}

\href{https://www.desmos.com/calculator/8k0cwumzso}{141: Exponential Growth Disk 1}

\end{exploration}

\begin{exploration}  \label{E5545636521}
Near the start of a pandemic $30$ people are infected with a virus. Twenty four hours later $40$ people are infected. 

Some time later, $300,000$ people are infected. How much longer will it take for an additional $100,000$ people to become infected? Assume the virus spreads exponentially. 
\end{exploration}

\begin{example} \label{Ed45345trt65tr5t}
\begin{enumerate}
\item Start with a piece of paper $0.01$ cm thick and fold it in half twenty times. About how thick is the resulting paper? Guess first, then compute an estimate.
\begin{hint}
\[
    2^{10} = 1024 \sim 10^3 .
\]
\end{hint}

\item How many more times would you need to fold the paper in half for it to be about $100$ km thick?
\end{enumerate}
\end{example}

\section{Growth Factors}
A population that grows \emph{linearly} increases at a constant absolute rate, perhaps at the rate of $20,000$ people/yr. 

A population that grows \emph{exponentially} increases at a constant \emph{relative} rate, perhaps at the rate of $20\%$/yr. To understand what this means, suppose the population was $P$ at the start of the year 2020. Then one year later the population would be $20\%$ greater, or 
\[
   P + 20\% P  = P + 0.2 P = 1.2P .
\]
We call 1.2 the \emph{annual growth factor}. To get the population one year from, we just multiply the current population by this factor of 1.2

\begin{question}  \label{Q00:ExponentialG}
A population grows exponentially, increasing by $25\%$ each year. Suppose the current population is $P$.

(a) What will the population be one year from now?
\[
\answer{1.25P}
\]
 
(b) What was the population one year ago?
\[
\answer{0.8P}
\]

(c) What will the population be three years from now?
\[
\answer{1.953125P}
\]

(d) What is the three-year growth factor?
\[
\answer{1.953125}
\]

(e) What is the relative change in the population over a three-year period?
\[
\text{The population increases by } \answer{95.3125}\% \text{ every three years.}
\]

\end{question}


\begin{question}  \label{Q1:ExponentialG}
Suppose a population decreases exponentially, decreasing by $20\%$ each year and that the current population is $P$. 

(a) What will the population be one year from now?  
\[
\answer{0.8P}
\]

(b) What is the annual growth factor?
\[
   \answer{0.8}
\]

(c) What was the population one year ago?
\[
\answer{1.25P}
\]

(d) What will the population be two years from now?
\[
\answer{0.64P}
\]

(e) What is the two-year growth factor?
\[
\answer{0.64}
\]

(f) What is the relative change in the population over a two-year period?
\[
\text{The population decreases by } \answer{36}\% \text{ every two years.}
\]

\end{question}

\begin{question}  \label{Q2:ExponentialG}
(a) Suppose a population increases by $20\%$ over a one-year period and then decreases by $20\%$ over the next year. What is the relative change in the population over the two-year period?
\[
   \text{The relative change in the population over the two-year period is }   \answer{-0.04} .
\]

\end{question}

\begin{question}  \label{Q3:ExponentialG}
A population increases by $40\%$ each year. Find the relative change in the population over a two-year period. 
\begin{multipleChoice}  
\choice{$80\%$}  
\choice[correct]{$96\%$}  
\choice{$20\%$}  
\end{multipleChoice}  
\end{question}




\section{Point-Slope vs. Point-Growth, Part 1}

\begin{example} \label{E567743344}
Suppose the population of a (small) colony of bacteria is $40,000$ at 2pm and $50,000$ at 3pm.

\begin{enumerate}

\item Suppose first that between 2pm and 8pm the population is a linear function of time.

\begin{enumerate}
\item What does it mean for the population to be a linear function of time? Explain in everday English. No math terms.

\item Find the population's growth rate.

The population increases at the constant rate of $\answer{10000}$ bac/hour.

\item Let 
\[
      P = f(t) , 2\leq t \leq 8 ,
\]
be a function that expresses the population in terms of the number of hours past noon.

\begin{enumerate}
\item Use common sense (do not rely on a formula) to evalute $f(8)$. Show the complete computation. Include units for every number. Explain your reasoning. Click on the arrow at the lower right for guidance.

\begin{expandable}
To find the population $f(8)$ at 8pm, we start with the population $f(2) = 40000$ at 2pm and add to that the \emph{change} in the population from 2pm to 8pm.

Because the length of the time interval from 2pm to 8pm is
\[
   \Delta t = \answer{8} \text{ hrs} - \answer{2}\text{ hrs} = \answer{6} \text{ hrs}
\]
and because the population increases at the constant rate of 
\[
  \frac{\Delta P}{\Delta t}  = \answer{10000} \text{ bac/hr} ,
\]
the change in population from 2pm to 8pm is
\[
  \Delta P = f(\answer{8}) - f(\answer{2}) = (\answer{10,000} \text{ bac/hr}) (\answer{6}\text{ hrs}) = \answer{60000}\text{ bac} . 
\]
So the population at 8pm is
\[
    f(8) = \answer{40000} + \answer{10000} ( 8 - \answer{2} ) = 100000
\]
bacteria.

\end{expandable}



\item Use the same reasoning to find an expression for the function $P=f(t)$. Explain this reasoning.
\end{enumerate}

\end{enumerate}

\item Next suppose that between 2pm and 8pm the population is an exponential function of time.

\begin{enumerate}
\item What does it mean for the population to be an exponential function of time? Explain in everday English. No math terms.

\item Compute the one-hour growth factor. What are its units?%At what rate does the population grow?

\item Interpret the meaning of the one-hour growth factor.

Every hour the population gets multipled by $\answer{1.25}$.

\pskip

Or, every hour the population increases by $\answer{25}\%$.


\item Let 
\[
      P = f(t) , 2\leq t \leq 8 ,
\]
be a function that expresses the population in terms of the number of hours past noon.

\begin{enumerate}
\item Use common sense (do not rely on a formula) to evalute $f(8)$. Show the complete computation. Include units for every number. Explain your reasoning. Click the arrow to the lower right for guidance.
\begin{expandable}
The key idea is to use the one-hour growth factor of $\answer{1.25}$. Because each hour the population gets mutiplied by the factor $\answer{1.25}$ and because the length of the time interval from 2pm to 8pm is
\[
   \Delta t = \answer{8} \text{ hrs} - \answer{2}\text{ hrs} = \answer{6} \text{ hrs} , 
\] 
we need only mutiply the population at 2pm by the factor $\answer{1.25}$ six times. So the population at 8pm is
\[
   f(8) = f(2) (\answer{1.25})^{\answer{6}} = \answer{40000} (\answer{1.25})^{\answer{6}} 
\]
bacteria.
 
\end{expandable}

\item Use the same reasoning to find an expression for the function $P=f(t)$. Explain this reasoning.
\end{enumerate}
\end{enumerate}

\end{enumerate}

\end{example}

\begin{example}  \label{Ex432234e3r}
Between 2pm and 8pm the population of a colony of bacteria decreases exponentially. The population is $50,000$ at 2pm and $40,000$ at 3pm.

\begin{enumerate}
\item Find the one-hour growth factor. Interpret its meaning.

Every hour the population gets multipled by $\answer{0.80}$.

\pskip

Or, every hour the population decreases by $\answer{20}\%$.

\item Find a function
\[
        P = f(t) , 2\leq t \leq 8 ,   
\] 
that expresses the population in terms of the number of hours past noon.

\item Explain the logic behind your function in part (b).

\item Check that your function is correct by evaluating $f(2)$ and $f(3)$.
\end{enumerate}
\end{example}


\begin{example}  \label{Ex4556511111}
Between 2pm and 10pm the population of a (small) colony of bacteria increases exponentially. There are 50,000 bacteria at 3pm and 60,000 bacteria at 4pm

\begin{enumerate}
\item Find the population one hour after there are 100,000 bacteria.

\item Find the population one hour before there are 100,000 bacteria.

\item What happens to the population every two hours?

\item Find the population two hours after there are 130,000 bacteria

\item Find a function 
\[
      P = f(t) \, , \, 2\leq t \leq 10 ,
\]
that expresses the population in terms of the number of hours past noon. Explain your reasoning.
\end{enumerate}
\end{example}

\section{Point-Slope vs. Point-Growth, Part 2}
\begin{example} \label{Ex578788766}
Between 2pm and 10pm the population of a (small) colony of bacteria increases exponentially. There are 40,000 bacteria at 2pm and 90,000 bacteria at 5pm.

\begin{enumerate}
\item Find the three-hour growth factor and interpret its meaning. Include units.

\item Let $g$ be the one-hour growth factor. Write and solve an equation to find the \emph{exact} value of $g$. No calculator.

\item Find a function 
\[
      P = f(t) \, , \, 2\leq t \leq 10 ,
\]
that expresses the population in terms of the number of hours past noon. Explain your reasoning.

\item Use your function to find the population at 3:30pm.

\item Suppose instead the population had increased at a constant rate. What would the population at 3:30 have been?
\end{enumerate}
\end{example}

\begin{example} \label{Exdf5tr454tt88766}
Between 2pm and 10pm the population of a (small) colony of bacteria decreases exponentially. There are 80,000 bacteria at 2pm and 70,000 bacteria at 3:30pm.

\begin{enumerate}
\item Find the 1.5-hour growth factor and interpret its meaning. Include units.

\item Let $g$ be the one-hour growth factor. Write and solve an equation to find the \emph{exact} value of $g$. No calculator.

\item Find a function 
\[
      P = f(t) \, , \, 2\leq t \leq 10 ,
\]
that expresses the population in terms of the number of hours past noon. Explain your reasoning.

\item Use your function to find the population at 10pm.

\end{enumerate}
\end{example}

\begin{example} \label{Ex5434232388766}
The population of a colony of bacteria is $P_0$ at time $t=t_0$ hours past noon and $P_1$ at time $t=t_1$ hours past noon.
Assume $\leq 0 t_0, t_1 \leq 12$.

\begin{enumerate}

\item Suppose the population changes at a constant rate between noon and midnight.

\begin{enumerate}
\item Find a function 
\[
      P = f(t) \, , \, 2\leq t \leq 12 ,
\]
that expresses the population in terms of the number of hours past noon. Explain your reasoning.

\item Input your function on Line 1 of the worksheet below by correcting the function there.

\begin{onlineOnly}
    \begin{center}
\desmos{vdwsk61zza}{900}{600}
\end{center}
\end{onlineOnly}

\href{https://www.desmos.com/calculator/vdwsk61zza}{141: Exp vs Linear Growth}

\item Check to see you are correct by dragging each of the two points $(t_0, P_0)$ and $(t_1, P_1)$.

\item Use your function to find a simplified expression for the population at time $t=(t_0+t_1)/2$.

\end{enumerate}


\item Suppose the population either increases or decreases exponentially between noon and midnight. 


\begin{enumerate}
\item Find a function 
\[
      P = g(t) \, , \, 2\leq t \leq 12 ,
\]
that expresses the population in terms of the number of hours past noon. Explain your reasoning.

\item Input your function on Line 2 of the worksheet below by correcting the function there.

\item Check to see you are correct by dragging each of the two points $(t_0, P_0)$ and $(t_1, P_1)$.

\item Use your function to find a simplified expression for the population at time $t=(t_0+t_1)/2$.

\end{enumerate}

\item Compare the populations at time $t=(t_0+t_1)/2$ in the linear and exponential models. Can you say for sure if one is always at least as great as the other? Explain. If so, write an inequality that expresses this fact.

\end{enumerate}
\end{example}




\section{Exercises}
\begin{question}   \label{Q4:ExponentialG}
The population of a colony of bacteria increases exponentially between 12pm and 6pm. The population is $400,000$ at 12:00pm and $440,000$ at 1:00pm. 

(a) What is the one-hour growth factor? Explain what this means.

(b) What happens to the population every hour?

(c) What is the three-hour growth factor? What happens to the population every three hours?

(d) Find a function 
\[
     P = f(t) \, , 0\leq t \leq 6 
\]
that expresses the population in terms of the number of hours past noon.

\end{question}


\begin{question}   \label{Q5:ExponentialG}
The population of a colony of bacteria increases exponentially between 12pm and 6pm. The population is $400,000$ at 1:00pm and $520,000$ at 4:00pm. 

(a) What is the three-hour growth factor?

(b) What happens to the population every three hours?

(c) What is the one-hour growth factor?

(d) Find a function 
\[
     P = f(t) \, , 0\leq t \leq 6 
\]
that expresses the population in terms of the number of hours past noon.

\end{question}



\begin{question}   \label{Q5:ExdsfderdflG}
The population of a colony of bacteria decreases exponentially between 12pm and 6pm. The population was $400,000$ at 1:00pm and $300,000$ at 2:00pm. 

(a) What is the one-hour growth factor?

(b) What happens to the population every hour?

(c) Find a function 
\[
     P = f(t) \, , 0\leq t \leq 6 
\]
that expresses the population in terms of the number of hours past noon.

(d) What was the population at noon?

\end{question}



\begin{question}   \label{Q6:ExponentialG}
The world population was approximately 1.5 billion in 1900 and 2.5 billion in 1950. Assume that between 1900 and 2023 the population grew exponentially.

(a) Describe what happened to the population every 50 years.

(b) Describe what happened to the population every year.

(c) What is the one-year growth factor? Give an exact value and then an approximate value to 4 decimal places.
\[
   \text{Exact growth factor }  \answer{(5/3)^{1/50}} \hskip 0.3 in  \text{Approximate growth factor } \answer{1.0103}
\]

(d) Use the exact one-year growth factor to find a function
\[
     P = f(t) \, , 0\leq t \leq 123 ,
\]
that expresses the population (measured in billions of people) in terms of the number of years past 1900.

\[
   f(t) = \answer{1.5 \left(\frac{5}{3}\right)^{t/50}}
\]

(e) Use your function to predict the world population at the end of 2023. How does this compare with the actual population?

\end{question}


\begin{question}   \label{Q88:ExponentialG}
Suppose that between sea level and an altitude of 10 km, the air pressure is an exponential function of altitude. Suppose further that the pressure is 14.7 pounds/(square inch) at sea level and 10.17 pounds/(square inch) at an altitude of 3 km.

(a) What happens to the air pressure when the altitude increases by 3 km?

(b) What happens to the air pressure when the altitude increases by 1 km?

(c) Find a function
\[
    P = f(h) \, , 0\leq h \leq 10,
\]
that expresses the pressure (measured in lbs/(square inch) in terms of the altitude (measured in kilometers). Enter the function below.
\[
      f(h) = \answer{14.7(10.17/14.7)^{h/3}} \, , 0\leq h \leq 10 .
\]

(d) Enter your function in Line 1 of the exploration below (ie. replace the incorrect function $f(h)=14.7-h$ with the correct one) and use the graph to estimate the altitude at which the pressure is 7 lbs/(square inch).


\pdfOnly{
Access Desmos interactives through the online version of this text at
 
\href{https://www.desmos.com/calculator/a8d1scpez0}.
}
 
\begin{onlineOnly}
    \begin{center}
\desmos{a8d1scpez0}{900}{600}
\end{center}
\end{onlineOnly}

\end{question}


\begin{question}   \label{Q22:ExponentialG}
(a) Suppose that a colony of bacteria has a population of 100,000 at 1pm and a population of 490,000 at 4pm. Find functions 
\[
   P = f(t) \, , 0\leq t \leq 10 ,
\]
and 
\[
   P = g(t) \, , 0\leq t \leq 10 ,
\]
that express the population (measured in thousands of bacteria) in terms of the number of hours past noon assuming the population grows a constant rate (the function $f$) and at a constant relative rate (the function $g$), respectively. Use point-slope for $f$. {\bf Explain the logic behind the expressions for each function.}

Enter these functions in the worksheet below. Compare their behaviors over the given domain.

\pdfOnly{
Access Desmos interactives through the online version of this text at
 
\href{https://www.desmos.com/calculator/bwine7dcmg}.
}
 
\begin{onlineOnly}
    \begin{center}
\desmos{bwine7dcmg}{900}{600}
\end{center}
\end{onlineOnly}

\pskip

(b) Compare the steps  you took in finding the functions $f$ and $g$ in part (a) in the following way:

i) Think of the first step in finding the linear function as finding the change in the population from 1pm to 4pm. How did you modify this step for the exponential function?

ii) The second step in finding the linear function is determining the average growth rate. How did you modify this step for the exponential function?

iii) The third step in finding the linear function (really two steps at once) is to add the change in the population from 1pm to time $t$ hours past noon to the population at 1pm. How did you modify this step for the exponential function?

iv) You used three arithmetical operations in finding the linear function - subtraction (Step 1), multiplication (Steps 2 and 3), and addition (Step 3). What operations replaced these in the exponential function?



(c) Repeat part (a) assuming the population is $a\times 10^3$ at 1pm and $b\times 10^3$ at 4pm. Use the parameters $a$ and $b$, not specific numbers. Then open the Sliders Folder in Line 1 by clicking the arrow at the left of the folder's name. Vary the sliders for $a$ and $b$ and describe how the graphs change. Be sure to consider a decreasing population.


(d) For part (c), find fully-simplified expressions (in terms of $a$ and $b$) for the populations at 2:30pm.

\end{question}


\begin{question} \label{Q23:ExponentialG}
(a) The function 
\[
   f(t) = P_0 (2^{t/33}) \, , -10\leq t \leq 50 ,
\]
expresses the population of a country in terms of the number of years past 1940. Describe precisely how the population grows. Do not use a calculator.

(b) The function 
\[
   m = m_0 \left( \frac{3}{4} \right)^{t/1300} \, , -10,000\leq t \leq 1000 ,
\]
expresses the mass (in kilograms) of a radiocative compound in terms of the number of years since 1900. Describe precisely how the mass decays. Do not use a calculator.

\end{question}



\begin{question}   \label{Q9:ExponentialG}
Suppose both the population and national debt of a small country grow exponentially. The population doubles every 20 years and the national debt doubles every 12 years. How long does it take the per-capita (ie. per-person) share of the national debt to double?
\[
    \answer{30} \text{ years}
\]
\end{question}



\begin{question} \label{Q24:ExponentialG}
The function 
\[
   P = f(t) \, , -40\leq t \leq 60 ,
\]
expresses the population of a small country (measured in hundreds of thousands of people) in terms of the number of years since 1900. Assume that the popluation increases between 1860 and 1960 and interpret the meaning of each of the following expressions.

(a) $f^{-1}(20)$

(b) $f(f^{-1}(20) + 30)$

(c) $f^{-1}(1.44 f(30))$

\pskip

Now assume that the population increases by $20\%$ every 10 years and the expressions in parts (b) and (c) above. Explain your reasoning. Do {\bf not} find an explicit expression for the inverse function nor refer to logarithmic functions in any way.


\end{question}


\begin{question}   \label{Q10:ExponentialG}
Returning to the scenario of Question 7, 

(a) find a function
\[
    P = g(T) \, , 0\leq T \leq 373 ,
\]
that expresses the population in terms of the number of years since 1650. 

{\it Hint:} First express $t$ (the number of years since 1900) in terms of $T$.

(b) Use your function from part (a) to find a function
\[
  A = g_1(T) \, , 0\leq T \leq 373 ,
\]
that expresses the land area (measured in billions of hectares) needed to feed the world population as a function of the number of years since 1650 (see the caption to Figure 10 below and assume the present (1970) productivity level).

(c) Use your function from part (a) to find functions
\[
  A = g_2(T)  \text{ and } A = g_3(T)\, , 0\leq T \leq 373 ,
\]
that express the land area (measured in billions of hectares) needed to feed the world population as a function of the number of years since 1650 at levels of productivity double and quadruple the 1970 level).

(d) Find a function
\[
    A = h(T) \, , 250\leq T \leq 373 ,
\]
that expresses the area (in billions of hectares) available for agriculture as a function of the number of years since 1650. Assume there were 3.2 billion acres available in 1900 and that each additional person added to the population after 1900 required 0.08 hectares of arable land for infrastructure that rendered that land unavailable for agriculture.

\pdfOnly{
Access Desmos interactives through the online version of this text at
 
\href{https://www.desmos.com/calculator/ovjygr01yy}.
}
 
\begin{onlineOnly}
    \begin{center}
\desmos{ovjygr01yy}{900}{600}
\end{center}
\end{onlineOnly}


\end{question}


\begin{question} \label{Q12:ExponentialG}

The table below shows popluations of four colonies of bacteria at certain times during the day. The time is measured in the number of hours past noon.
\begin{center}
  \begin{tabular}{ | l | c | c |c|c| }
    \hline 
    Time & $P_1$ & $P_2$ & $P_3$ & $P_4$\\ \hline
    0 & 200,000  & 200,000 & 200,000 & 200,000\\ \hline
    1 & 230,000 & 230,000 & 210,000 & \\ \hline
    2 &   & 264,500 & & 216,320\\ \hline
    3 & 290,000 & &231,525 & 223,530\\  \hline
   % 4 &  \\  & \hline
   % \hline
  \end{tabular}
\end{center}

Which of the populations appear to be growing

(a) linearly?

(b) exponentially?

(c) neither linearly nor exponentially?

\pskip

Justify your assertions with computations along with complete explanations that describe what you are computing. 

Then compute the missing entries for those populations that appear to be growing linearly or exponentially. Explain your work.

\end{question}




\begin{example} \label{E1:ExpGrowth}
Suppose the population of a colony of bacteria changes exponentially between noon and 6pm. For each of the following scenarios, do the following:

i) Describe the growth factor over the appropriate period.

ii) Describe the relative change in the population over that period.

iii) Find a function
\[
       P = f(t) \, , 0\leq t \leq 6 ,
\]
that expresses the population (measured in hundreds of thousands of bacteria) in terms of the number of hours past noon.

iv) Find the population at 6pm.

\pskip

(a) The population is $400,000$ at noon and $450,000$ at 1pm.

(b) The population is $400,000$ at noon and $300,000$ at 1pm.

(c) The population is $400,000$ at noon and $500,000$ at 2pm.

(d) The population is $400,000$ at noon and $500,000$ at 2:30pm.

(e)  The population is $400,000$ at noon and $300,000$ at 2:20pm.

(f)  The population is $400,000$ at 1:00pm and $500,000$ at 3:00pm.

\end{example}


\end{document}