\documentclass{ximera}
\title{Linear Functions CW}

\newcommand{\pskip}{\vskip 0.1 in}

\begin{document}
\begin{abstract}
The point-slope equation of a line.
\end{abstract}
\maketitle


\begin{question}  \label{QLer3r3r3}
Between 12:15pm and 1:00pm a balloon descends at a constant rate of $37$ ft/min.

The balloon is at an altitude of $3235$ feet at 12:34 pm.

\begin{enumerate}
\item Use common sense to find the altitude of the balloon at 12:55. Explain your reasoning. Include units for each number in your computation.

\item Using the same logic as in part (a), find an expression for the function 
\[
  h = f(t) \, , \, 15\leq t \leq 60 .
\]
that give the balloon's height (in feet) at time $t$ minutes past noon.

\begin{hint}
During the 
\[
    \Delta t = t - 34
\]
minutes from 12:34pm to time $t$ minutes past noon, the change in altitude is
\begin{align*}
    \Delta h &= \left( -37 \frac{\text{ ft}}{\text{min}} \right) \left( \Delta t \text{ min} \right) \\
                &= -37 (t-34) \text{ ft}. 
\end{align*}
Adding this change to the height $f(34) = 3235$ feet of the balloon at 12:34pm give the balloon's  height (measured in feet)
\begin{align*}
   h   &= f(t) \\
        &= f(34) + -37 (t-34) \\
        &= 3235 - 37(t-34) \, , \, 15\leq t \leq 60,
\end{align*}
at time $t$ minutes past noon. 
\end{hint}

\item Use logic, \emph{not} your function, to determine when the balloon is at an altitude of $2828$ feet.

\item Use your function from part (b) to determine when the balloon is at an altitude of $2828$ feet. Your algebraic solution should be identical to your logical solution from part (c).
\end{enumerate}

\end{question}


\begin{question} \label{QKdfder33}
The graph of a linear function $y=f(x)$ passes through the points $(-4,5)$ and $(7,-3)$.

\begin{enumerate}
\item Sketch a reasonably accurate graph of the function $y=f(x)$.

\item Use point-slope to find an expression for the function $y=f(x)$. Explain your logic. Part of your explanation should include a sentence that begins with the phrase

 \emph{A point $P$ (different from $(4,5)$) with coordinates $(x,y)$ lies on the line through $(-4,5)$ with slope ?? if and only if ...}

\item Use your function from part (b) to solve the inequality
\[
     f(x) < 29 .
\]
Write your answer as a solution set.

\item Solve the inequality
\[
      f(x) < f(29) .
\]
Write your answer as a solution set.
\end{enumerate}
\end{question}

\begin{question} \label{QLfefrre3}
The equation 
\[
  25p + 7q = 300 \, ,\, 5 \leq p \leq 12 ,
\]
relates the average number of burgers $q$ a food truck sells per day in terms of the price $p$ (measured in dollars/burger).

What happens to the average number of burgers/day sold if they decrease the price by $\$0.77$/burger? Be specific. Try to solve this problem in general \emph{without} using any specific numbers.
 
\begin{hint}
Suppose the average number of burgers sold/day changes from $q$ to $q+\Delta q$ as the price changes from $p$ to $p +\Delta p$ dollars/burger. Then
\[
    25 (p+ \Delta p) + 7 (q + \Delta q) = 300.
\]
Using the distributive property of multiplication over addition and the commutative property of addition gives
\[
     (25p + 7q) + (25 \Delta p + 7 \Delta q) = 300.
\]
But since
\[
    25p + 7q = 300 ,
\]
it must be that
\[
       25 \Delta p + 7 \Delta q = 0.
\]
Solving for $\Delta q$, we get
\[
       \Delta q = -\frac{25}{7} \Delta p.
\]
So if they decrease the price by $\$0.77$/burger (ie. if $\Delta p = -\$0.77$/burger), then
\begin{align*}
     \Delta q &= -\left(  \frac{25\text{ burgers/day }}{7 \text{ dollars/burger}}   \right) \left( - 0.77 \text{ dollars/burger} \right) \\
                  &= 2.75 \text{ burgers/day} 
\end{align*}
and they would sell an average of $2.75$ more burgers/day.
\end{hint}

\end{question}

\begin{question}  \label{Qldfe3rf}
The function
\[
    g = f(v) \, , \, 50\leq v\leq 80,
\]
expresses the gas mileage (in miles/gal) of a car in terms of its speed (in miles/hour).
%expresses the average number of burgers a diner sells per day in terms of the price (measured in dollars/burger).

\begin{enumerate}
\item Increasing the speed by $5\%$ decreases the gas mileage by $2$ miles/gal. What equation would you solve to determine the original speed? 

\item Decreasing the speed by $5$ miles/hour increases the gas mileage by $10\%$. What equation would you solve to determine the original speed?  

\item Suppose now 
\[
  g = f(v) = 40 - \frac{3}{5}(v-52) \, , \, 50\leq v\leq 80.
\]
Answer the two questions above.
%the price at which a $\$2$/burger increase will decrease the average number of burgers sold per day by $25\%$?

\item  
\end{enumerate}

\end{question}

\begin{question}  \label{Qeeetl4gg}
Between speeds of $55$ miles/hour and $90$ miles/hour, the gas mileage of a Corvette is a linear function of its speed. The car gets $35$ miles/gallon at a speed of $55$ miles/hour and $20$ miles/gallon at a speed of $90$ miles/hour.

\begin{enumerate}

\item Find a function $v=f(G)$ that expresses the speed of the Corvette (in miles/hour) in terms of its gas mileage (in miles/gallon). Start by sketching a graph of the function. Label the axes with the appropriate variable names and units.

\item Use set notation to state the domain and range of $f$.

\item What are the units of the slope of your line in part (a)? Interpret its meaning.

\item At what speed does the car get $30$ miles/gal?

\item What is the gas mileage at a speed of $80$  miles/hour?

\item What happens to the gas mileage when the speed of the car decreases by $5$ miles/hour?

\item How can you increase the gas mileage by $4$ miles/gal?

\item Find a function $r=h(G)$ that expresses the rate at which the car burns gas (in gal/hour) in terms of its gas mileage. State the domain and range of $h$ in set notation.

\item Simplify the units of the slope in part (c). What does this suggest about its meaning? Would that be a correct interpretation?

\item Drag the slider $u$ in the worksheet below to approximate the rate (in miles/hr) at which the Corvette burns gas at a speed of $86$ miles/hour. Then use the appropriate function to compute the exact rate.

\item Drag the slider $u$ in the worksheet below to approximate the speed at which the Corvette burns gas at the rate of $2$ gal/hour. Then use the appropriate function to compute the exact rate.

\begin{onlineOnly}
    \begin{center}
\desmos{2fq4txcouq}{900}{600}
\end{center}
\end{onlineOnly}

\href{https://www.desmos.com/calculator/2fq4txcouq}{141: Corvette}

\item Find a function $G=w(r)$ that expresses the gas mileage (in miles/hour) of the Corvette in terms of the rate (in gal/hour) at which it burns gas. State the domain and range $w$ in set notation.

\end{enumerate}
\end{question}


\end{document}